---
title: "Income- k-means"
author: "Karol Orozco"
date: "2023-04-10"
output: html_document
---

## Setup
```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(tidytext)
library(caret)
library(fastDummies)
library(randomForest)
# https://www.openml.org/d/1590

raw_income = read_csv("https://raw.githubusercontent.com/chanks06/ml-model3/main/datasets/openml_1590.csv", na=c("?"))

income = read_csv("https://raw.githubusercontent.com/chanks06/ml-model3/main/datasets/openml_1590.csv", na=c("?")) %>%
  drop_na() %>%
  mutate(income_above_50k = class==">50K") %>%
  select(-class) %>%
  dummy_cols(remove_selected_columns = T)

```

```{r}
head(income)
```

10 pts k-means

* discussion for choosing number of clusters
* analysis of cluster centers
* bivariate chart(s) against meaningful variables and/or analysis of density plots
* meaningful interpretation / discussion of conclusions 

10 pts supervised learning

* feature engineering / selection, whether with PCA or otherwise
* interpretation of variable importance, coefficients if applicable
* justification of choice of metric
* discussion of choice or tuning of hyperparameters, if any
* meaningful discussion of predictive power and conclusions from model